HO unification from object language to meta language

### 2024-04-05 paper/main.tex
small fix

\noindent
As per \cref{inv:linklhs} the term on the left is a variable, and its
right counterpart is the
term in \maybeeta, and that term has the following property:

@@

% In the examples above, the first term is a \maybeeta since $A_x$ can
% reduce to $x$ by setting $A_x = \lambda x.x$, 
% the second one is not a \maybeeta since it exists no substitution
% for $A_x$ such that $A_x$ reduces to $x$ and $x$ is not free in the subterm $f \appsep x$.
\noindent
The first two examples are easy, and show how a unification variable can expose
or erase a variable and turn the resulting term in an $\eta$-expansion or not.\\
The third example shows that when a variable occurs outside the scope of a unification
variable it cannot be erased and can hence prevent a term from being an $\eta$-expansion.\\
The last example shows the recursive nature of the check we need to implement.
The term starts with a spine of two lambdas hence the whole term
is in \maybeeta iff the inner term $\lambda y.f\appsep (A \appsep x) \appsep (B \appsep y \appsep x)$
is in \maybeeta itself. If it is, it could $\eta$-contract to
$f \appsep (A \appsep x)$ making $\lambda x.f \appsep (A \appsep x)$ a potential
$\eta$-expansion.\\

We can now define more formally how \maybeeta terms are detected together with
its auxiliary functions:

\noteE[inline]{the compiler must do beta normal!!!!}
\newcommand{\reduceto}{\emph{may-contract-to}}
\begin{definition}[\reduceto]
  A term $s$ \reduceto{} a name $x$ if there exists a
  substitution $\rho$ such that $\rho s \Eo{} x$.
\end{definition}

\begin{lemma}\label{lem:reduceto}
A $\beta$-normal term $s = \lambda x_1 \ldots x_n.t$
%, where $x, x_1\ldots x_n$ can occur in $t$, 
\reduceto{} $x$ only if one of the following three conditions holds:
\begin{enumerate}
  \item $n = 0$ and $t = x$;
  \item $t$ is the application of $x$ to a list
     of terms $l$ and each $l_i$ \reduceto{} $x_i$
     (e.g. $\lambda x_1 \ldots x_n.x \appsep x_1 \ldots x_n \Eo{} x$) ;
  \item $t$ is a unification variable with scope
    $W$, and for any $v \in \{ x, x_1 \ldots x_n \}$,
    there exists a $w_i \in W$, such that $w_i$ \reduceto{} $v$
    (if $n = 0$ this is equivalent to $x \in W$).
\end{enumerate}
\end{lemma}
\begin{proof}[Proof sketch]
Since our terms are in $\beta$-normal form there is only
one rule that can play a role (namely \ref{clause:eta1}),
hence if the term $s$ is not exactly $x$ (case 1)
it can only be an $\eta$-expansion of $x$, or a unification
variable that can be assigned to $x$, or a combination of both.
If $s$ begins with a lambda, then the lambda can only disappear by $\eta$
contraction. In that case the term $t$ under the spine of binders for
$x_1\ldots x_n$ can either be $x$ itself applied to terms that can
\reduceto{} these variables (case 2), or a unification variable
that can be assigned to that application (case 3).
\end{proof}

\noindent
Note that this condition does not require the term to be in \llambda.
\noteE[inline]{Is this relevant}

\newcommand{\occursrigid}{\emph{occurs-rigidly}}
\newcommand{\occurrigid}{\emph{occur-rigidly}}
\begin{definition}[\occursrigid]\label{def:occrigid}
  A name $x$ \occursrigid{} in a $\beta$-normal term $t$, if $\forall \rho, x \in
  \subterm{\rho t}$
\end{definition}
In other words $x$ \emph{\occursrigid} in $t$ if it occurs in $t$
outside of the scope of unification variables since an instantiation
is allowed to discard $x$ from the scope of the unification variable.
Note that $\eta$-contraction cannot make $x$ disappear, since the
variables being erased by $\eta$-contraction are locally bound inside $t$.

@@

\newcommand{\testmaybeeta}{\emph{maybe-eta}\xspace}
\begin{definition}[\testmaybeeta]\label{def:testmaybeeta}
  Given a $\beta$-normal term
  $s = \lambda x_1 \ldots x_n.t$, \testmaybeeta{} $s$ holds if any
  of the following holds:
  \begin{enumerate}
    \item $t$ is a constant or a variable applied to the arguments
      $l_1 \ldots l_m$ such that 
      $m \geq n$ and for every $i$ such that $1 \leq i \leq m-n$
      the term  $l_i$
      \reduceto{} $x_i$, and
      no $x_i$ \occursrigid{} in $l_1 \ldots l_{m-n-1}$;
    \item $t$ is a
      unification variable with scope $W$ and
      for each $x_i$ there exists a $w_j \in W$ such that $w_j$
      \reduceto{} $x_i$.
  \end{enumerate}
\end{definition}
\begin{lemma}[\maybeeta detection]\label{lem:maybeeta}
  If $t$ is a $\beta$-normal term and \testmaybeeta{} $t$ holds,
  then $t \in \maybeeta$.
\end{lemma}
\begin{proof}[Proof sketch]
Follows from \cref{def:occrigid} and \cref{lem:reduceto}
\end{proof}

@@

\noindent
The rule triggers when the input term \elpiIn{flam F} is in
\maybeeta. It compiles it to \elpiIn{lam F1} but puts the fresh
variable \elpiIn{A} in its place. The variable sees all the names free in
\elpiIn{lam F1}. The critical part of this rule is the creation of the \linketa,
which relates the variable \elpiIn{A} with \elpiIn{lam F1}.
This link clearly validates \cref{inv:linklhs}.

@@

\begin{proof}[Proof sketch]
  By contradiction, suppose that the rule above triggered and that
  the \rhs of the link is $\lambda x.\lambda y.t_{xy}$.
  If $\testmaybeeta{}~\lambda y.t_{xy}$ holds the recursive call to
  \elpiIn{comp} (made by \elpiIn{comp-lam}) must have put a fresh variable
  in its place, so this case is impossible.
  Otherwise, if $\testmaybeeta{}~\lambda y.t_{xy}$ does not hold, also
  $\testmaybeeta{}~\lambda x.\lambda y.t_{xy}$ does not hold, contradicting
  the assumption that the rule triggered.
\end{proof}

@@

\linketa are meant to delay the unification of ``problematic'' terms until
we know for sure if the head lambda has to be $\eta$-contracted or not.

>>>>

\noindent
As per \cref{inv:linklhs} the term on the left is a variable, and its
right counterpart is the
term in \maybeeta. That term has the following property:

@@

% In the examples above, the first term is a \maybeeta since $A_x$ can
% reduce to $x$ by setting $A_x = \lambda x.x$, 
% the second one is not a \maybeeta since it exists no substitution
% for $A_x$ such that $A_x$ reduces to $x$ and $x$ is not free in the subterm $f \appsep x$.
\noindent
The first two examples are easy, and show how a unification variable can expose
or erase a variable in their scope and turn the resulting term in an $\eta$-expansion or not.\\
The third example shows that when a variable occurs outside the scope of a unification
variable it cannot be erased and can hence prevent a term from being an $\eta$-expansion.\\
The last example shows the recursive nature of the check we need to implement.
The term starts with a spine of two lambdas hence the whole term
is in \maybeeta iff the inner term $\lambda y.f\appsep (A \appsep x) \appsep (B \appsep y \appsep x)$
is in \maybeeta itself. If it is, it could $\eta$-contract to
$f \appsep (A \appsep x)$ making $\lambda x.f \appsep (A \appsep x)$ a potential
$\eta$-expansion.\\
We can now define more formally how \maybeeta terms are detected together with
its auxiliary functions:

\newcommand{\reduceto}{\emph{may-contract-to}}
\begin{definition}[\reduceto]
  A $\beta$-normal term $s$ \reduceto{} a name $x$ if there exists a
  substitution $\rho$ such that $\rho s \Eo{} x$.
\end{definition}

\begin{lemma}\label{lem:reduceto}
A $\beta$-normal term $s = \lambda x_1 \ldots x_n.t$
%, where $x, x_1\ldots x_n$ can occur in $t$, 
\reduceto{} $x$ only if one of the following three conditions holds:
\begin{enumerate}
  \item $n = 0$ and $t = x$;
  \item $t$ is the application of $x$ to a list
     of terms $l$ and each $l_i$ \reduceto{} $x_i$
     (e.g. $\lambda x_1 \ldots x_n.x \appsep x_1 \ldots x_n \Eo{} x$) ;
  \item $t$ is a unification variable with scope
    $W$, and for any $v \in \{ x, x_1 \ldots x_n \}$,
    there exists a $w_i \in W$, such that $w_i$ \reduceto{} $v$
    (if $n = 0$ this is equivalent to $x \in W$).
\end{enumerate}
\end{lemma}
\begin{proof}[Proof sketch]
Since our terms are in $\beta$-normal form there is only
one rule that can play a role (namely \ref{clause:eta1}),
hence if the term $s$ is not exactly $x$ (case 1)
it can only be an $\eta$-expansion of $x$, or a unification
variable that can be assigned to $x$, or a combination of both.
If $s$ begins with a lambda, then the lambda can only disappear by $\eta$
contraction. In that case the term $t$ is under the spine of binders
$x_1\ldots x_n$, $t$ can either be $x$ applied to terms that can
\reduceto{} these variables (case 2), or a unification variable
that can be assigned to that application (case 3).
\end{proof}

% \noindent
% Note that this condition does not require the term to be in \llambda.
% \noteE[inline]{Is this relevant}

\newcommand{\occursrigid}{\emph{occurs-rigidly}}
\newcommand{\occurrigid}{\emph{occur-rigidly}}
\begin{definition}[\occursrigid]\label{def:occrigid}
  A name $x$ \occursrigid{} in a $\beta$-normal term $t$, if $\forall \rho, x \in
  \subterm{\rho t}$
\end{definition}

\noteD{Not too clear?}
In other words $x$ \emph{\occursrigid} in $t$ if it occurs in $t$
outside of the scope of unification variables since an instantiation
is allowed to discard $x$ from the scope of the unification variable.
Note that $\eta$-contraction cannot make $x$ disappear, since the
variables being erased by $\eta$-contraction are locally bound inside $t$.

@@

\newcommand{\testmaybeeta}{\emph{maybe-eta}\xspace}
\begin{definition}[\testmaybeeta]\label{def:testmaybeeta}
  Given a $\beta$-normal term
  $s = \lambda x_1 \ldots x_n.t$, \testmaybeeta{} $s$ holds if any
  of the following holds:
  \begin{enumerate}
    \item $t$ is a constant or a name applied to the arguments
      $l_1 \ldots l_m$ such that 
      $m \geq n$ and for every $i$ such that $m - n < i \leq m$
      the term  $l_i$
      \reduceto{} $x_i$, and
      no $x_i$ \occursrigid{} in $l_1 \ldots l_{m-n}$;
    \item $t$ is a
      unification variable with scope $W$ and
      for each $x_i$ there exists a $w_j \in W$ such that $w_j$
      \reduceto{} $x_i$.
  \end{enumerate}
\end{definition}
\begin{lemma}[\maybeeta detection]\label{lem:maybeeta}
  If $t$ is a $\beta$-normal term and \testmaybeeta{} $t$ holds,
  then $t \in \maybeeta$.
\end{lemma}
\begin{proof}[Proof sketch]
Follows from \cref{def:occrigid} and \cref{lem:reduceto}
\end{proof}

@@

\noindent
The rule triggers when the input term \elpiIn{flam F} is in
\maybeeta. It compiles \elpiIn{flam F} to \elpiIn{lam F1} but puts the fresh
variable \elpiIn{A} in its place. This variable sees all the names free in
\elpiIn{lam F1}. The critical part of this rule is the creation of the \linketa,
which relates the variable \elpiIn{A} with \elpiIn{lam F1}.
This link clearly validates \cref{inv:linklhs}.

@@

\begin{proof}[Proof sketch]
  By contradiction, suppose that the rule above is triggered and that
  the \rhs of the link is $\lambda x.\lambda y.t_{xy}$.
  If $\testmaybeeta{}~\lambda y.t_{xy}$ holds the recursive call to
  \elpiIn{comp} (made by \elpiIn{comp-lam}) must have put a fresh variable
  in its place, so this case is impossible.
  Otherwise, if $\testmaybeeta{}~\lambda y.t_{xy}$ does not hold, also
  $\testmaybeeta{}~\lambda x.\lambda y.t_{xy}$ does not hold, contradicting
  the assumption that the rule triggered.
\end{proof}

@@

\linketa are meant to delay the unification of ``problematic'' terms until
we know for sure if the term has to be $\eta$-contracted or not.
